Synthesizing Quotient Lenses

### 2018-06-24 papers/DifferenceLenses/main.tex
Made mandatory changes

The functions $\delta_S$ and $\delta_V$ will depend only on $S$ and $V$, and will be designed to measure how similar two sources (or views) are to each other. More specifically, given a regular expression $R$ and strings $s, s' \in \mathcal{L}(R)$, we will define a number $\delta_R(s, s')$ inductively on $R$, with $0 \leq \delta_R(s, s') \leq 1$, and a high value of $\delta_R(s, s')$ indicates that $s$ and $s'$ are similar---in particular, if $s = s'$, then $\delta_R(s, s') = 1$, and if $s \neq s'$ and either $s = \varepsilon$ or $s' = \varepsilon$, then $\delta_R(s, s') = 0$. We will also expand our definition of the regular expression $R^*$, having three different variants $R^*_{POS}$, $R^*_{NC}$ and $R^*_{MED}$, where the differences are computed using positional, non-crossing and min-edit distance alignment strategies respectively, so that for all $s, s' \in \mathcal{L}(R^*)$, $\delta_{R^*_{MED}}(s, s') \geq \delta_{R^*_{NC}}(s, s') \geq \delta_{R^*_{MED}}(s, s')$.
\subsection{Difference Functions on Regular Languages}
\begin{definition}
Given a finite alphabet $\Sigma$, a {\em difference regular expression} is any expression generated by the following grammar,
$$R := s \; | \; R_1 \cdot R_2 \; | \; R_1 \cup R_2 \; | \; R^*_{POS} \; | \; R^*_{NC} \; | \; R^*_{MED}$$
where $s$ ranges over $\Sigma^*$ and $R$ ranges over regular expressions. The languages of $R^*_{POS}$, $R^*_{NC}$ and $R^*_{MED}$ are all the same, namely $\mathcal{L}(R^*)$.
\end{definition}
\begin{definition}
Given an unambiguous difference regular expression $R$, define $\delta_R : \mathcal{L}(R) \times \mathcal{L}(R) \longrightarrow \mathbb{R}^{\geq 0}$ inductively as follows:
\begin{align*}
\delta_s(t, t') &= 1\\
\delta_{R_1 \cdot R_2}(s_1 \cdot s_2, s'_1 \cdot s'_2) &= \frac{1}{2}(\delta_{R_1}(s_1, s'_1) + \delta_{R_2}(s_2, s'_2))\\
\delta_{R_1 \cup R_2}(s, s') &= 
\begin{cases}
\delta_{R_1}(s, s') & \text{if }s, s' \in \mathcal{L}(R_1) \\
\delta_{R_2}(s, s') & \text{if }s, s' \in \mathcal{L}(R_2)\\
0 & \text{otherwise }
\end{cases}\\
\delta_{R^*_{POS}}(s_1 \cdot \ldots \cdot s_m, t_1 \cdot \ldots \cdot t_n) &= 
\begin{cases}
0 & \text{if } m=0 \text{ or } n=0\\
\max\limits_{\{\varphi : [m] \longrightarrow [n] \; | \; \varphi(i) = i\}}\frac{1}{n}\sum\limits_{i=1}^{m}\ \delta_R(s_i, t_{\varphi(i)})& \text{if } m \leq n\\
\max\limits_{\{\varphi : [n] \longrightarrow [m] \; | \; \varphi(i) = i\}}\frac{1}{m}\sum\limits_{i=1}^{m}\ \delta_R(t_i, s_{\varphi(i)})& \text{otherwise}
\end{cases}\\
\delta_{R^*_{NC}}(s_1 \cdot \ldots \cdot s_m, t_1 \cdot \ldots \cdot t_n) &= 
\begin{cases}
0 & \text{if } m=0 \text{ or } n=0\\
\max\limits_{\{\varphi : [m] \longrightarrow [n] \; | \; \varphi \text{ is strictly increasing}\}}\frac{1}{n}\sum\limits_{i=1}^{m}\ \delta_R(s_i, t_{\varphi(i)})& \text{if } m \leq n\\
\max\limits_{\{\varphi : [n] \longrightarrow [m] \; | \; \varphi \text{ is strictly increasing}\}}\frac{1}{m}\sum\limits_{i=1}^{m}\ \delta_R(t_i, s_{\varphi(i)})& \text{otherwise}
\end{cases}\\
\delta_{R^*_{MED}}(s_1 \cdot \ldots \cdot s_m, t_1 \cdot \ldots \cdot t_n) &= 
\begin{cases}
0 & \text{if } m=0 \text{ or } n=0\\
\max\limits_{\{\varphi : [m] \longrightarrow [n] \; | \; \varphi \text{ is injective}\}}\frac{1}{n}\sum\limits_{i=1}^{m}\ \delta_R(s_i, t_{\varphi(i)})& \text{if } m \leq n\\
\max\limits_{\{\varphi : [n] \longrightarrow [m] \; | \; \varphi \text{ is injective}\}}\frac{1}{m}\sum\limits_{i=1}^{m}\ \delta_R(t_i, s_{\varphi(i)})& \text{otherwise}
\end{cases}
\end{align*}
\end{definition}
\begin{fact}
For any regular expression $R$ and $s, s' \in \mathcal{L}(R)$
\begin{enumerate}
\item
$0 \leq \delta_R(s, s') \leq 1$.
\item
$\delta_R(s, s) = 1$.
\item
If $s \neq s'$ and either $s = \varepsilon$ or $s' = \varepsilon$, then $\delta_R(s, s') = 0$
\item
For all $s, s' \in \mathcal{L}(R^*)$, $\delta_{R^*_{MED}}(s, s') \geq \delta_{R^*_{NC}}(s, s') \geq \delta_{R^*_{MED}}(s, s')$.
\item
$\delta_R(s, s') = \delta_R(s', s)$.
\end{enumerate}

@@

let rec $\strunfilter \; S \; xs \; ys$ = 
match $xs, ys$ with
$|$ $\varepsilon$,_ -> $ys$
$|$ $x \cdot xs'$,_ when $x \in S$ -> $x \cdot (\strunfilter \; S \; xs' \; ys)$
$|$ $x \cdot xs',\varepsilon$ when $x \not \in S$ -> $\strunfilter \; R \; xs' \; \varepsilon$
$|$ $x \cdot xs',y \cdot ys'$ when $x \not \in S$ -> $y \cdot (\strunfilter \; R \; xs' \; ys')$
\end{lstlisting}
Given a source $s \in (R \cup S)^*$, the lens $\filter \; R \; S$ removes all elements that match $S$ in $s$ in the \get direction and restores them in the \pput direction.
\begin{claim}
Let $R$ and $S$ be difference regular expressions with $\mathcal{L}(R) \cap \mathcal{L}(S) = \varnothing$ and $(R \cup S)^{!*}$. Then $\filter_{POS} \; R \; S : (R \sqcup S)^*_{MED} \Leftrightarrow R^*_{MED}$ is a difference lens.
\end{claim}
\begin{proof}
We need to show that
$$
\delta_{(R \cup S)^*_{MED}} (\strunfilter \; S \; s \; v, s) \geq \delta_{R^*_{MED}}(v, \strfilter \; R \; s)$$
Observe that if $s = s_1 \cdot \ldots cdot s_m$ with $s_{i_1}, \ldots, s_{i_p} \in R$ and $s_{j_1}, \ldots, s_{j_p} \in S$, then $\strfilter \; R \; s = s_{i_1} \cdot \ldots \cdot s_{i_n}$. Assume that $v = v_1 \cdot \ldots v_{n'}$ and assume without loss of generality that $n \leq n'$. Then since $\delta_{R^*_{MED}}(v, \strfilter \; R \; s) \leq 1$, it follows that
$$
\delta_{(R \cup S)^*_{MED}} (\strunfilter \; S \; s \; v, s) = \frac{m + \delta_{R^*_{MED}}(v, \strfilter \; R \; s) * n'}{m+n'} \geq \delta_{R^*_{MED}}(v, \strfilter \; R \; s)
$$
which is what we wanted to show.
\end{proof}
\subsubsection{Merge}
If $R$ is a difference regular expression with $R \cdot^! R$, then define the \mergelens lens $\mergelens \; R : R \cdot R \Leftrightarrow R$ as follows:
(\mergelens \; R).\get \; (s \cdot s') &= s\\
(\mergelens \; R).\pput \; (s \cdot s') \; t &= 
t \cdot t & \text{if }s = s'\\
t \cdot s' & \text{otherwise}
(\mergelens \; R).\create s &= s \cdot s
\begin{claim}
If $R$ is a difference regular expression with $R \cdot^! R$, then $\mergelens \; R : R \cdot R \Leftrightarrow R$ is a difference lens.
\end{claim}
\begin{proof}
We need to show that
$$\delta_{R \cdot R}((\mergelens \; R).\pput \; (s \cdot s') \; t, (s \cdot s')) \geq \delta_R(t, s)$$
If $s = s'$, then we need to show that
$$\delta_{R \cdot R}(t \cdot t, s \cdot s) \geq \delta_R(t, s)$$
which follows from the fact that
If $s \neq s'$, then we need to show that
$$\delta_{R \cdot R}(t \cdot t, s \cdot s) = \frac{1}{2}\left(\delta_R(t, s) + \delta_R(t, s)\right) = \delta_R(t, s)$$
$$\delta_{R \cdot R}(t \cdot s', s \cdot s') = \frac{1}{2}\left(\delta_R(t, s) + \delta_R(s', s')\right) = \frac{1}{2}\left(\delta_R(t, s) + 1\right) \geq \delta_R(t, s)$$
which follows from the fact that $\delta_R(t, s) \leq 1$.
\end{proof}
\iffalse
\section{Matching Lenses}
A {\em basic lens} $\ell : S \xLeftrightarrow{C} V$ is a 4-tuple of functions
$\ell.\get : S \longrightarrow V$, $\ell.\res : S \longrightarrow C$,
$\ell.\pput : V \longrightarrow C \longrightarrow S$ and $\ell.\create : V
\longrightarrow S$ satisfying the following properties:
\ell.\get \; (\ell.\pput \; v \; c) &= v \tag{PUTGET}\\
\ell.\pput \; (\ell.\get \; s) \; (\ell.\res \; s) &= s \tag{GETPUT}
Idea: First remove matched parts to leave unmatched skeleton. Next remove
unmatched parts to leave matched skeleton and then flatten.
\section{Relational Lenses}
We define a {\em relation} $R$ of type $((T_1, \ldots, T_n), P, F)$ to be a
multiset $R$ where $T_i, P$ are sets satisfying $R \subseteq P \subseteq T_1
\times \ldots \times T_n$, $F \subseteq [n] \times [n]$, and if $i, j \in F$,
then for all $r, r' \in R$, if $r_i = r'_i$ then $r_j = r'_j$. We interpret the
$T_i$ to be a set of {\em attributes}, $R$ to be a set of {\em records} with
$dom(r) = (T_1, \ldots, T_n)$ for all $r \in R$, $P$ to be a {\em predicate}
such that the records in $R$ satisfy $P$, and $F$ to be a set of {\em
functional dependencies} on the attributes of the records in $R$.

>>>>

The functions $\delta_S$ and $\delta_V$ will depend only on $S$ and $V$, and will be designed to measure how similar two sources (or views) are to each other. More specifically, given a regular expression $R$ and strings $s, s' \in \mathcal{L}(R)$, we will define a number $\delta_R(s, s')$ inductively on $R$, with $0 \leq \delta_R(s, s') \leq 1$, and a high value of $\delta_R(s, s')$ indicates that $s$ and $s'$ are similar---in particular, if $s = s'$, then $\delta_R(s, s') = 1$, and if $s \neq s'$ and either $s = \varepsilon$ or $s' = \varepsilon$, then $\delta_R(s, s') = 0$. We will also expand our definition of the regular expression $R^*$, having three different variants $R^*_{POS}$, $R^*_{NC}$ and $R^*_{MED}$, where the differences are computed using positional, non-crossing and min-edit distance alignment strategies respectively, so that for all $s, s' \in \mathcal{L}(R^*)$, $\delta_{R^*_{MED}}(s, s') \geq \delta_{R^*_{NC}}(s, s') \geq \delta_{R^*_{MED}}(s, s')$.
\subsection{Difference Functions on Regular Languages}
Given a finite alphabet $\Sigma$, a {\em difference regular expression} is any expression generated by the following grammar,
$$R := s \; | \; R_1 \cdot R_2 \; | \; R_1 \cup R_2 \; | \; R^*_{POS} \; | \; R^*_{NC} \; | \; R^*_{MED}$$
where $s$ ranges over $\Sigma^*$ and $R$ ranges over regular expressions. The languages of $R^*_{POS}$, $R^*_{NC}$ and $R^*_{MED}$ are all the same, namely $\mathcal{L}(R^*)$.
Given an unambiguous difference regular expression $R$, define $\delta_R : \mathcal{L}(R) \times \mathcal{L}(R) \longrightarrow \mathbb{R}^{\geq 0}$ inductively as follows:
\delta_s(t, t') &= 1\\
\delta_{R_1 \cdot R_2}(s_1 \cdot s_2, s'_1 \cdot s'_2) &= \frac{1}{2}(\delta_{R_1}(s_1, s'_1) + \delta_{R_2}(s_2, s'_2))\\
\delta_{R_1 \cup R_2}(s, s') &= 
\delta_{R_1}(s, s') & \text{if }s, s' \in \mathcal{L}(R_1) \\
\delta_{R_2}(s, s') & \text{if }s, s' \in \mathcal{L}(R_2)\\
0 & \text{otherwise }
\delta_{R^*_{POS}}(s_1 \cdot \ldots \cdot s_m, t_1 \cdot \ldots \cdot t_n) &= 
1 & \text{if } m = 0, n = 0\\ 
0 & \text{if } m = 0, n \neq 0\\
0 & \text{if } m \neq 0, n = 0\\\max\limits_{\{\varphi : [m] \longrightarrow [n] \; | \; \varphi(i) = i\}}\frac{1}{n}\sum\limits_{i=1}^{m}\ \delta_R(s_i, t_{\varphi(i)})& \text{if } m \leq n\\
\max\limits_{\{\varphi : [n] \longrightarrow [m] \; | \; \varphi(i) = i\}}\frac{1}{m}\sum\limits_{i=1}^{m}\ \delta_R(t_i, s_{\varphi(i)})& \text{otherwise}
\delta_{R^*_{NC}}(s_1 \cdot \ldots \cdot s_m, t_1 \cdot \ldots \cdot t_n) &= 
1 & \text{if } m = 0, n = 0\\ 
0 & \text{if } m = 0, n \neq 0\\
0 & \text{if } m \neq 0, n = 0\\
\max\limits_{\{\varphi : [m] \longrightarrow [n] \; | \; \varphi \text{ is strictly increasing}\}}\frac{1}{n}\sum\limits_{i=1}^{m}\ \delta_R(s_i, t_{\varphi(i)})& \text{if } m \leq n\\
\max\limits_{\{\varphi : [n] \longrightarrow [m] \; | \; \varphi \text{ is strictly increasing}\}}\frac{1}{m}\sum\limits_{i=1}^{m}\ \delta_R(t_i, s_{\varphi(i)})& \text{otherwise}
\delta_{R^*_{MED}}(s_1 \cdot \ldots \cdot s_m, t_1 \cdot \ldots \cdot t_n) &= 
1 & \text{if } m = 0, n = 0\\ 
0 & \text{if } m = 0, n \neq 0\\
0 & \text{if } m \neq 0, n = 0\\\max\limits_{\{\varphi : [m] \longrightarrow [n] \; | \; \varphi \text{ is injective}\}}\frac{1}{n}\sum\limits_{i=1}^{m}\ \delta_R(s_i, t_{\varphi(i)})& \text{if } m \leq n\\
\max\limits_{\{\varphi : [n] \longrightarrow [m] \; | \; \varphi \text{ is injective}\}}\frac{1}{m}\sum\limits_{i=1}^{m}\ \delta_R(t_i, s_{\varphi(i)})& \text{otherwise}
\end{cases}
\begin{fact}
For any regular expression $R$ and $s, s' \in \mathcal{L}(R)$
\begin{enumerate}
\item
$0 \leq \delta_R(s, s') \leq 1$.
\item
$\delta_R(s, s) = 1$.
\item
If $s \neq s'$ and either $s = \varepsilon$ or $s' = \varepsilon$, then $\delta_R(s, s') = 0$
\item
For all $s, s' \in \mathcal{L}(R^*)$, $\delta_{R^*_{MED}}(s, s') \geq \delta_{R^*_{NC}}(s, s') \geq \delta_{R^*_{MED}}(s, s')$.
\item
$\delta_R(s, s') = \delta_R(s', s)$.
\end{enumerate}

@@

let rec $\strunfilter \; S \; xs \; ys$ = 
match $xs, ys$ with
$|$ $\varepsilon$,_ -> $ys$
$|$ $x \cdot xs'$,_ when $x \in S$ -> $x \cdot (\strunfilter \; S \; xs' \; ys)$
$|$ $x \cdot xs',\varepsilon$ when $x \not \in S$ -> $\strunfilter \; R \; xs' \; \varepsilon$
$|$ $x \cdot xs',y \cdot ys'$ when $x \not \in S$ -> $y \cdot (\strunfilter \; R \; xs' \; ys')$
\end{lstlisting}
Given a source $s \in (R \cup S)^*$, the lens $\filter \; R \; S$ removes all elements that match $S$ in $s$ in the \get direction and restores them in the \pput direction.
\begin{claim}
Let $R$ and $S$ be difference regular expressions with $\mathcal{L}(R) \cap \mathcal{L}(S) = \varnothing$ and $(R \cup S)^{!*}$. Then $\filter_{POS} \; R \; S : (R \sqcup S)^*_{MED} \Leftrightarrow R^*_{MED}$ is a difference lens.
\end{claim}
\begin{proof}
We need to show that
$$
\delta_{(R \cup S)^*_{MED}} (\strunfilter \; S \; s \; v, s) \geq \delta_{R^*_{MED}}(v, \strfilter \; R \; s)$$
Observe that if $s = s_1 \cdot \ldots cdot s_m$ with $s_{i_1}, \ldots, s_{i_p} \in R$ and $s_{j_1}, \ldots, s_{j_p} \in S$, then $\strfilter \; R \; s = s_{i_1} \cdot \ldots \cdot s_{i_n}$. Assume that $v = v_1 \cdot \ldots v_{n'}$ and assume without loss of generality that $n \leq n'$. Then since $\delta_{R^*_{MED}}(v, \strfilter \; R \; s) \leq 1$, it follows that
$$
\delta_{(R \cup S)^*_{MED}} (\strunfilter \; S \; s \; v, s) = \frac{m + \delta_{R^*_{MED}}(v, \strfilter \; R \; s) * n'}{m+n'} \geq \delta_{R^*_{MED}}(v, \strfilter \; R \; s)
$$
which is what we wanted to show.
\end{proof}
\subsubsection{Merge}
If $R$ is a difference regular expression with $R \cdot^! R$, then define the \mergelens lens $\mergelens \; R : R \cdot R \Leftrightarrow R$ as follows:
\begin{align*}
(\mergelens \; R).\get \; (s \cdot s') &= s\\
(\mergelens \; R).\pput \; (s \cdot s') \; t &= 
\begin{cases}
t \cdot t & \text{if }s = s'\\
t \cdot s' & \text{otherwise}
\end{cases}\\
(\mergelens \; R).\create s &= s \cdot s
\end{align*}
\begin{claim}
If $R$ is a difference regular expression with $R \cdot^! R$, then $\mergelens \; R : R \cdot R \Leftrightarrow R$ is a difference lens.
\end{claim}
\begin{proof}
We need to show that
$$\delta_{R \cdot R}((\mergelens \; R).\pput \; (s \cdot s') \; t, (s \cdot s')) \geq \delta_R(t, s)$$
If $s = s'$, then we need to show that
$$\delta_{R \cdot R}(t \cdot t, s \cdot s) \geq \delta_R(t, s)$$
which follows from the fact that
If $s \neq s'$, then we need to show that
$$\delta_{R \cdot R}(t \cdot t, s \cdot s) = \frac{1}{2}\left(\delta_R(t, s) + \delta_R(t, s)\right) = \delta_R(t, s)$$
$$\delta_{R \cdot R}(t \cdot s', s \cdot s') = \frac{1}{2}\left(\delta_R(t, s) + \delta_R(s', s')\right) = \frac{1}{2}\left(\delta_R(t, s) + 1\right) \geq \delta_R(t, s)$$
which follows from the fact that $\delta_R(t, s) \leq 1$.
\end{proof}
\subsection{Matching Lenses}
Difference lenses provide three iteration combinators for dealing with the issue of {\em local alignment}. However, if the ``chunks'' that a programmer wishes to align do not occur underneath a star regular expression, then the difference lenses combinators will not align them. Consider for example the lens:
\begin{lstlisting}
let k : lens = (copy [A-Z]) . ([a-z]<->"")
let l : lens = k . (copy "," . k)*
\end{lstlisting}
If we try to put the view \lstinline|"Z,Y,W,X"| into the source \lstinline|"Xx,Yy,Zz"|, then using a min-edit distance iteration lens, we will get the result \lstinline|"Zx,Yy,Wz,Xa"|, because the view element $``Z"$, being out outside the star, will be put into $``Xx"$, and the iteration lens will align the chunks remaining in the star. What is needed in this example is a lens which performs a ``global alignment'' strategy.

Barbosa's {\em matching lenses} \cite{barbosa2010matching} do just this, by allowing a programmer to mark every chunk that they wish to align anywhere in a lens term. The guarantee that matching lenses give is that alignment used to match view chunks is the same alignment that is used to match source chunks.

\iffalse
\section{Relational Lenses}
We define a {\em relation} $R$ of type $((T_1, \ldots, T_n), P, F)$ to be a
multiset $R$ where $T_i, P$ are sets satisfying $R \subseteq P \subseteq T_1
\times \ldots \times T_n$, $F \subseteq [n] \times [n]$, and if $i, j \in F$,
then for all $r, r' \in R$, if $r_i = r'_i$ then $r_j = r'_j$. We interpret the
$T_i$ to be a set of {\em attributes}, $R$ to be a set of {\em records} with
$dom(r) = (T_1, \ldots, T_n)$ for all $r \in R$, $P$ to be a {\em predicate}
such that the records in $R$ satisfy $P$, and $F$ to be a set of {\em
functional dependencies} on the attributes of the records in $R$.

### 2018-06-24 papers/QuotientLenses/main.tex
Made mandatory changes

Strong unambiguity is necessary in the premises of the regular QRE combinators
because it ensures that the canonizing function of a QRE is well-defined.
For example, consider the QRE \lstinline{"a"$^*$ . (collapse "a"$^*$ $\to$ "a")}. The
behavior of this QRE is not well-defined since the string
\cd{"aaa"} can be canonized to any of \cd{"a"}, \cd{"aa"}, or \cd{"aaa"}
depending on how \cd{"aaa"} is parsed. We also require that the regular
combinators applied to the kernels of QREs be unambiguous since the underlying
bijective string lens of a QRE lens operates on kernels, and bijective string lenses
impose the same unambiguity restrictions so that they too are well defined as
functions.

@@

To evaluate our QRE implementation, we adapted 39 lens synthesis tasks
from the benchmark suite of the original Optician system. These
benchmarks are a combination of custom benchmarks, benchmarks derived
from FlashFill~\cite{flashfill}, and benchmarks derived from
Augeas~\cite{augeas2}.  In these 39 benchmarks, 10 of
the data formats had to be modified to work with the bijectivity
constraints that Optician required. For instance, when one
representation permits whitespace where the other does not, we
modified the original version of the benchmark to allow more whitespace,
thereby restoring bijectivity (but altering the data format). With the
new QRE support, we were able to remove these alterations.  This
experience alone suggests that QREs make the lens development process
more flexible.

@@

Figure~\ref{fig:asts} shows each of these measures for the 10
non-bijective problems in the benchmark suite.  On
average, \canonizeAndSpecSize{} used~44\% more AST nodes
than \QRESize{}, requiring an average of~237 more AST nodes. On 
average, \LensAndSpecSize{} used~193\% more AST nodes than \QRESize{}, requiring an
average of~1267 more AST nodes. These figures suggest that introducing QREs saves
programmers significant effort compared to both Optician and basic
Boomerang.

>>>>

These unambiguity constraints can be a little fiddly in practice. As a simple example of this, consider writing a regular expression for comma-separated lists of strings. Our first impulse might be to write it as,
\begin{lstlisting}
CSL = anychar+ . ("," . anychar+)*
\end{lstlisting}
(where . is concatenation), but this regular expression is ambiguous, as \lstinline|anychar| is a big union of a of single characters including comma.

While the unambiguity constraints consequently appear to compromise compositionality of QREs, they can usually be circumvented by making a small changes to the offending regular expression: for instance, in the preceding example, then we need to write,
\begin{lstlisting}
CSL = anycharexceptcomma+ . ("," . anycharexceptcomma+)*
\end{lstlisting}
(assuming \lstinline|anycharexceptcomma| denotes a big union of every single character string except comma).

Unambiguity constraints are necessary because they ensure that the canonizing function of a QRE is well-defined.
For example, consider the QRE \lstinline|"a"$^*$ . (collapse "a"$^*$ $\to$ "a")|. The
behavior of this QRE is not well-defined since the string
\cd{"aaa"} can be canonized to any of \cd{"a"}, \cd{"aa"}, or \cd{"aaa"}
depending on how \cd{"aaa"} is parsed. The unambiguity constraints are also applied to the kernels of QREs since the underlying
bijective string lens of a QRE lens operates on kernels, and bijective string lenses
impose the same unambiguity restrictions so that they too are well defined as
functions.

@@

To evaluate our QRE implementation, we adapted 39 lens synthesis tasks
from the benchmark suite of the original Optician system. These
benchmarks are a combination of custom benchmarks, benchmarks derived
from FlashFill~\cite{flashfill}, and benchmarks derived from
Augeas~\cite{augeas2} (we also experimented using our QRE implementation to synthesize quotient lenses between XML, RDF and and JSON formats using data from data.gov database; the data consisted of census statistics, demographic statistics, wage comparosion data, and crime index data).  In these 39 benchmarks, 10 of
the data formats had to be modified to work with the bijectivity
constraints that Optician required. For instance, when one
representation permits whitespace where the other does not, we
modified the original version of the benchmark to allow more whitespace,
thereby restoring bijectivity (but altering the data format). With the
new QRE support, we were able to remove these alterations.  This
experience alone suggests that QREs make the lens development process
more flexible.

@@

Figure~\ref{fig:asts} shows each of these measures for the 10
non-bijective problems in the benchmark suite.  On
average (using a geometric mean), \canonizeAndSpecSize{} used~38.5\% more AST nodes
than \QRESize{}, requiring an average of~214 more AST nodes. On 
average, \LensAndSpecSize{} used~180\% more AST nodes than \QRESize{}, requiring an
average of~998 more AST nodes. These figures suggest that introducing QREs saves
programmers significant effort compared to both Optician and basic
Boomerang.
